---
title: "Synchro Full Dataset: Indicator Species"
author: "Emily Wapman"
format: html
editor: visual
---

### Import data and read libraries

```{r}
#| warning = FALSE
library(readr)
library(phyloseq)
library(here)
library(tidyverse)
library(taxize)
library(dplyr)
library(marmap)

sample_data_raw <- read_csv(here("shiny_dashboard/data/raw", "Synchro_fish_mammals_full.csv"), skip = 1)

#metadata_raw <- read_csv(here("shiny_dashboard/data/raw", "CTD_metadata_joined_for_R_10082026.csv"))

metadata_raw <- read_csv(here("shiny_dashboard/data/raw", "CTD_metadata_joined_for_R_18082026.csv"))

```

### Create phyloseq object

```{r}
#| warning = FALSE

# Create OTU table
otu_table <- sample_data_raw |> 
  select(-c(2:40)) |> # remove columns 2:40 (Keep Feature id and sample names)
  as.data.frame() # convert data.frame to matrix object class

rownames(otu_table) <- otu_table[,1] # make the Feature id column the row names
otu_table <- otu_table[,-1]

otu_phylo <- phyloseq::otu_table(otu_table, taxa_are_rows = TRUE) # create phyloseq otu object

# Create Taxa table
taxa_table <- sample_data_raw %>%
  select(c(1, 5, 10:15)) %>% # Keep 5 (preferred sp. for reporting) and 10-15 (Domain, etc.)
  as.matrix() # convert to matrix object

rownames(taxa_table) <- taxa_table[,1]
taxa_table <- taxa_table[,-1]

taxa_phylo <- phyloseq::tax_table(taxa_table) # create phyloseq object for taxa

# Create phyloseq object (OTU + Taxa)
physeq_otu_taxa <- phyloseq(otu_table = otu_phylo, tax_table = taxa_phylo)

# Add metadata to phyloseq object

# Extract sample names from phyloseq object (creates separate dataframe w/ just sample names)
physeq_sample_names <- data.frame(sample = sample_names(physeq_otu_taxa))

# Expand metadata by joining with physeq samples - only want ones with data
metadata_expanded <- metadata_raw |> 
  mutate(sample = SampleID) |> 
  right_join(physeq_sample_names, by = "sample") |> 
  dplyr::select(sample, everything()) |> 
  as.data.frame()

# Make the row names the sample names (first col) then drop the first col
rownames(metadata_expanded) <- metadata_expanded[,1]
metadata_expanded  <- metadata_expanded[,-1]

# Transform metadata to phyloseq object
metadata_expanded_phylo  <- phyloseq::sample_data(metadata_expanded)

# check if metadata is compatible with the phyloseq data
setdiff(sample_names(physeq_otu_taxa), rownames(metadata_expanded_phylo))
setdiff(rownames(metadata_expanded_phylo), sample_names(physeq_otu_taxa))

# merge phyloseq metadata with phyloseq object physeq_SP1_12S
physeq_synchro_full <- merge_phyloseq(physeq_otu_taxa, metadata_expanded_phylo)

```

### Removal of singletons/spurious OTUs based on a proportional threshold (0.001%)/sample

```{r}
#| warning = FALSE

# Filter low reads: 0.001 threshold (0.1%)
filter_rare_taxa <- function(physeq, threshold = 0.00001) {
  # threshold: minimum relative abundance (0.0001 = 0.01%)
  
  # Transform to relative abundance
  physeq_rel <- transform_sample_counts(physeq, function(x) x / sum(x))
  
  # Keep taxa that have at least the threshold abundance in at least one sample
  taxa_to_keep <- apply(otu_table(physeq_rel), 1, function(x) any(x >= threshold))
  
  # Filter original phyloseq object
  physeq_filtered <- prune_taxa(taxa_to_keep, physeq)
  
  cat(paste("Original taxa:", ntaxa(physeq), "\n"))
  cat(paste("Filtered taxa:", ntaxa(physeq_filtered), "\n"))
  cat(paste("Removed:", ntaxa(physeq) - ntaxa(physeq_filtered), "taxa\n"))
  
  return(physeq_filtered)
}

# Apply different threshold
ps_filtered_0.001 <- filter_rare_taxa(physeq_synchro_full, threshold = 0.00001)  # 0.001%  (Shelton et al 2016)
```

### List of what was removed after filtering

```{r}
#| warning = FALSE

# List of what was removed
removed_esv_list <- setdiff(taxa_names(physeq_synchro_full), taxa_names(ps_filtered_0.001))

# Extract taxonomy for removed ESVs

removed_tax <- as.data.frame(phyloseq::tax_table(physeq_synchro_full)) %>%
  tibble::rownames_to_column("ESV") %>%
  dplyr::filter(ESV %in% removed_esv_list)

# Extract OTU counts for removed ESVs
removed_otu <- as.data.frame(phyloseq::otu_table(physeq_synchro_full)) %>%
  tibble::rownames_to_column("ESV") %>%
  dplyr::filter(ESV %in% removed_esv_list)

# Merge and calculate total reads
removed_tax <- removed_tax %>%
  left_join(removed_otu, by = "ESV") %>%
  mutate(total_reads = rowSums(select(., where(is.numeric))))

# Select final columns
removed_tax_final <- removed_tax %>%
  select(ESV, Family, Genus, preferred_species_for_reporting, total_reads)

# View results
print(head(removed_tax_final))
print(paste("Number of removed ESVs:", nrow(removed_tax_final)))
print(paste("Total reads removed:", sum(removed_tax_final$total_reads)))
```

### Check that minimum abundance for each ASV is not below 10

```{r}
#| warning = FALSE
# Newest phyloseq object is ps_filtered_0.001

#taxa_sums looks at otu table totals across all samples per asv -- see if any are less than 10

# Get total abundance per ASV (sum across all samples)
asv_totals <- taxa_sums(ps_filtered_0.001)

# Check if any are below 10
any(asv_totals < 10)  # TRUE 

# How many are below 10?
sum(asv_totals < 10) # 279

# Remove ASV's less than 10 

ps_filtered_10 <- prune_taxa(taxa_sums(ps_filtered_0.001) >= 10, ps_filtered_0.001)


```

### Remove OTU's in negative controls proportionally from samples

```{r}
#| warning = FALSE

# Get negative controls (blanks)
controls <- subset_samples(ps_filtered_10, sample.type == "blank")

# calculate total read sum of all negative controls
total_nc_reads <- sum(taxa_sums(controls))

# Get cumulative read count for each ASV in controls
cummZOTU3 <- as.data.frame(taxa_sums(controls))
colnames(cummZOTU3)[1] <- "total_reads"
cummZOTU3$ZOTU <- rownames(cummZOTU3)

# Calculate relative abundance threshold for each ZOTU
cummZOTU3$threshold <- cummZOTU3$total_reads / total_nc_reads

# Get eDNA samples
data_samples_physeq_filtered <- subset_samples(ps_filtered_10, sample.type == "eDNA")

# Put names of negatives into a list
cont_negnames <- taxa_names(controls)

# Subset eDNA samples that contain negative controls
data_phyl_neg <- prune_taxa(cont_negnames, data_samples_physeq_filtered)

## Detach the otu table from the phyloseq object because we could not figure out how to overwrite the otu_table (make dataframe for easier manipulation)
table_detach3 <- as.data.frame(otu_table(data_phyl_neg)@.Data) # make otu table a data frame
sums_detach3 <- colSums(table_detach3) # Get the total number of reads per sample

## make sure it's the right order
table_detach3 <- table_detach3[match(cummZOTU3$ZOTU, rownames(table_detach3)), ]
## save sums of the sample in a vector

##Calculate relative abundance matrix
rel_abundance <- sweep(table_detach3, 2, sums_detach3, FUN = "/")

##Create threshold matrix for comparison
thresholds <- matrix(cummZOTU3$threshold,
                     nrow = nrow(table_detach3),
                     ncol = ncol(table_detach3),
                     byrow = FALSE)


## Filter out reads below contamination threshold
table_detach3[rel_abundance <= thresholds] <- 0

##Convert back to matrix and replace in phyloseq object
table_detach3 <- as.matrix(table_detach3)
otu_table(data_phyl_neg) <- otu_table(table_detach3, taxa_are_rows = TRUE)

# Store cleaned object
data_phyl_clean <- data_phyl_neg

#  Check remaining reads
sum(taxa_sums(ps_filtered_10))
sum(taxa_sums(data_phyl_clean))

sum(taxa_sums(ps_filtered_0.001)) - sum(taxa_sums(data_phyl_clean))

# New phyloseq object is data_phyl_clean


```

### Husbandry Removal

```{r}
#| warning = FALSE
#| 
ps_decontam <- subset_taxa(data_phyl_clean, !(preferred_species_for_reporting %in% c("Gallus gallus", "Meleagris gallopavo", "Homo sapiens", "Canis lupus familiaris", "Sus scrofa", "Bos taurus", "Mus mus musculus", "Felis catus", "Rhamdia cf. jequitinhonha NS 3056B")))

ntaxa(data_phyl_clean)
ntaxa(ps_decontam)

#ps_decontam is new phyloseq object
```

### Fix taxonomy: mammals

```{r}
#| warning = FALSE
# View unidentified taxa to locate marine mammals
# Extract taxonomy table
tax_df_decontam <- as.data.frame(phyloseq::tax_table(ps_decontam))

# Find rows where all are NA and put preferred species call in a list
unidentified <- tax_df_decontam |> 
  dplyr::filter(is.na(Domain) & is.na(Phylum) & is.na(Class) & is.na(Order) & is.na(Genus)) 
 
# Get mammal names
mammals <- unique(unidentified$preferred_species_for_reporting)

# Fill in Domain, Phylum, Class, Order, Family, Genus

tax_df_mammal_fix <- tax_df_decontam %>%
  mutate(
    Domain = case_when(
      preferred_species_for_reporting %in% mammals ~ "Eukaryota",
      TRUE ~ Domain
    ),
    Phylum = case_when(
      preferred_species_for_reporting %in% mammals ~ "Chordata",
      TRUE ~ Phylum
    ),
    Class = case_when(
      preferred_species_for_reporting %in% mammals ~ "Mammalia",
      TRUE ~ Class
    ),
    Order = case_when(
      preferred_species_for_reporting %in% c("Balaenoptera acutorostrata", "Balaenoptera musculus", 
                     "Balaenoptera physalus", "Megaptera novaeangliae") ~ "Mysticeti",
      preferred_species_for_reporting %in% c("Orcinus orca", "Phocoena phocoena") ~ "Odontoceti",
      preferred_species_for_reporting %in% c("Phoca vitulina", "Mirounga angustirostris", "Enhydra lutris") ~ "Carnivora",
      TRUE ~ Order
    ),
    Family = case_when(
      preferred_species_for_reporting %in% c("Balaenoptera acutorostrata", "Balaenoptera musculus", 
                     "Balaenoptera physalus", "Megaptera novaeangliae") ~ "Balaenopteridae",
      preferred_species_for_reporting == "Orcinus orca" ~ "Delphinidae",
      preferred_species_for_reporting == "Phocoena phocoena" ~ "Phocoenidae",
      preferred_species_for_reporting %in% c("Phoca vitulina", "Mirounga angustirostris") ~ "Phocidae",
      preferred_species_for_reporting == "Enhydra lutris" ~ "Mustelidae",
      TRUE ~ Family
    ),
    Genus = case_when(
      preferred_species_for_reporting %in% c("Balaenoptera acutorostrata", "Balaenoptera musculus", 
                     "Balaenoptera physalus") ~ "Balaenoptera",
      preferred_species_for_reporting == "Megaptera novaeangliae" ~ "Megaptera",
      preferred_species_for_reporting == "Orcinus orca" ~ "Orcinus",
      preferred_species_for_reporting == "Phocoena phocoena" ~ "Phocoena",
      preferred_species_for_reporting == "Phoca vitulina" ~ "Phoca",
      preferred_species_for_reporting == "Mirounga angustirostris" ~ "Mirounga",
      preferred_species_for_reporting == "Enhydra lutris" ~ "Enhydra",
      TRUE ~ Genus
    )
  )


```

### Fix taxonomy: Find cases with a species call but empty taxonomy columns

```{r}
#| warning = FALSE

# Find taxa with a species call but other taxonomy is NA
tax_df_prob_sp <- tax_df_mammal_fix |> 
  dplyr::filter(!is.na(preferred_species_for_reporting) & 
           (is.na(Domain) | is.na(Phylum) | is.na(Class) |
           is.na(Order) | is.na(Family) | is.na(Genus)))

# If genus is NA, fill with the first word of the species call
tax_df_filled <- tax_df_mammal_fix %>%
  mutate(
    Genus = if_else(
      (is.na(Genus) | Genus == "") & !is.na(preferred_species_for_reporting) & preferred_species_for_reporting != "",
      word(preferred_species_for_reporting, 1),
      Genus
    )
  )

# Look at remaining NA's
tax_df_prob_sp <- tax_df_filled |> 
  dplyr::filter(!is.na(preferred_species_for_reporting) & 
           (is.na(Domain) | is.na(Phylum) | is.na(Class) |
           is.na(Order) | is.na(Family) | is.na(Genus)))

# Citharichthys sordidus is missing family 

tax_df_filled <- tax_df_filled |> 
  mutate(
  Family = case_when(
      preferred_species_for_reporting == "Citharichthys sordidus" ~ "Cyclopsettidae",
      TRUE ~ Family
    )
  )

# Cymatogaster aggregata is missing order and family 

tax_df_filled <- tax_df_filled |> 
  mutate(
  Order = case_when(
      preferred_species_for_reporting == "Cymatogaster aggregata" ~ "Perciformes",
      TRUE ~ Order
    ),

  Family = case_when(
      preferred_species_for_reporting == "Cymatogaster aggregata" ~ "Embiotocidae",
      TRUE ~ Family
    ),
  )

# Put taxa_filled back into phyloseq object

# Convert to matrix
tax_matrix <- as.matrix(tax_df_filled) # convert back to matrix

# Put back into phyloseq
phyloseq::tax_table(ps_decontam) <- phyloseq::tax_table(tax_matrix)

# Verify it worked
phyloseq::tax_table(ps_decontam)[1:5, ]

```

Save psdecontam data as RDS file -- all detections

```{r}
# Extract taxonomy table (just need # of detections for now)
taxa_full_df <- as.data.frame(tax_table(ps_decontam))


saveRDS(taxa_full_df, here("shiny_dashboard", "data", "processed", "full_taxa.rds"))

# Extract full sample data

sample_data_df <- as.data.frame(sample_data(ps_decontam))

# Save it
saveRDS(sample_data_df, here("shiny_dashboard", "data", "processed", "sample_metadata.rds"))

# extract full otu data
otu_data <- as.data.frame(otu_table(ps_decontam))

# Save it for future use
saveRDS(otu_data, here("shiny_dashboard", "data", "processed", "otu_table.rds"))
```

### Import indicator species data

```{r}
library(janitor)
indicators <- read_csv(here("shiny_dashboard/data/raw", "Sanctuary_indicator_species_MBNMS.csv")) |> 
  clean_names()

```

### Find which indicator species are in our dataset

```{r}
# Join so that remaining are in both datasets 
# phyloseq object is ps_decontam

# Extract taxonomy table 
tax_df <- as.data.frame(tax_table(ps_decontam)) |> 
  rownames_to_column("ASV_ID")

# Indicator table has two habitat types for some species. Collapse to avoid duplicate ASV's
indicators_collapsed <- indicators |> 
  group_by(scientific_name) |> 
  summarise(
    habitat = paste(unique(habitat), collapse = "; "),  # Combine unique habitats
    across(-habitat, first)  # Keep other columns
  )

# Separate indicator taxa assigned to species level -- two words or does not end in spp
indicators_collapsed <- indicators_collapsed %>%
  mutate(
    is_species = str_detect(scientific_name, "^[A-Z][a-z]+ [a-z]+$") & 
                 !str_detect(scientific_name, "spp")
  )
  
# Split the species and higher classifications into two groups
ind_sp <- indicators_collapsed |> 
  filter(is_species == TRUE)

ind_higher <- indicators_collapsed |> 
  filter(is_species == FALSE) |>
  mutate(clean_name = str_remove(scientific_name, " spp\\.$")) # remove spp for joining

# First join to species dataframe 
tax_joined <- tax_df |> 
  left_join(
    ind_sp %>% select(scientific_name, 
                      habitat_sp = habitat,      
                      common_sp = common_name),
    by = c("preferred_species_for_reporting" = "scientific_name")
  )

# Join to higher taxonomy dataframe 

# Try Genus
tax_joined <- tax_joined %>%
  left_join(
    ind_higher %>% select(taxon = clean_name, 
                          habitat_genus = habitat, 
                          common_genus = common_name),
    by = c("Genus" = "taxon")
  )

# Try Family
tax_joined <- tax_joined %>%
  left_join(
    ind_higher %>% select(taxon = clean_name, 
                          habitat_family = habitat, 
                          common_family = common_name),
    by = c("Family" = "taxon")
  )

# Try Order
tax_joined <- tax_joined %>%
  left_join(
    ind_higher %>% select(taxon = clean_name, 
                          habitat_order = habitat, 
                          common_order = common_name),
    by = c("Order" = "taxon")
  )

# Try Class
tax_joined <- tax_joined %>%
  left_join(
    ind_higher %>% select(taxon = clean_name, 
                          habitat_class = habitat, 
                          common_class = common_name),
    by = c("Class" = "taxon")
  )


#Combine columns (priority: species > genus > family > order > class)
# Combine columns class ends up empty I think, so make it go to NA if empty to avoid problems
tax_joined <- tax_joined %>%
  mutate(
    habitat = coalesce(habitat_sp, habitat_genus, habitat_family, habitat_order, # coalesce returns the first non NA value from left to right
                       if("habitat_class" %in% names(.)) habitat_class else NA_character_),
    common_name_final = coalesce(common_sp, common_genus, common_family, common_order,
                                 if("common_class" %in% names(.)) common_class else NA_character_),
    match_level = case_when(
      !is.na(habitat_sp) ~ "species",
      !is.na(habitat_genus) ~ "genus",
      !is.na(habitat_family) ~ "family",
      !is.na(habitat_order) ~ "order",
      !is.na(habitat_class) ~ "class", 
      TRUE ~ "no_match"
    )
  ) %>%
  select(-any_of(c("habitat_sp", "habitat_genus", "habitat_family", # remove these added columns if they exist
                   "habitat_order", "habitat_class",
                   "common_sp", "common_genus", "common_family", 
                   "common_order", "common_class")))



# Filter to only those with matches
tax_joined_indicators <- tax_joined %>%
  filter(match_level != "no_match")

# Put ASVs back as rownames to put back in phyloseq object
tax_joined_indicators <- tax_joined_indicators |> 
  column_to_rownames("ASV_ID")

# Convert back to matrix and update phyloseq object
tax_table(ps_decontam) <- tax_table(as.matrix(tax_joined_indicators))


```

### Get data ready for plotting and save as RDS object for shiny app

```{r}
# First get otu table from phyloseq
otu_df <- otu_table(ps_decontam) |> 
  as.data.frame() |> 
  rownames_to_column("ASV_ID")

# Pivot longer so that this can be combined with the tax/metadata
otu_long <- otu_df |> 
  pivot_longer(
    cols = -ASV_ID,
    names_to = "sample_id",
    values_to = "abundance"
  ) |> 
  filter(abundance > 0)

# Join with tax table and metadata 
metadata_df <- as.data.frame(sample_data(ps_decontam)) 
tax_df <- as.data.frame(tax_table(ps_decontam)) |> 
  rownames_to_column("ASV_ID")

shiny_data <- otu_long |> 
  left_join(tax_df, by = "ASV_ID") |> 
  left_join(metadata_df, by = c("sample_id" = "SampleID")) # CHANGED FROM SAMPLE.NAME

# Add a column with the correct name to use for mapping -- lowest assignment
shiny_data <- shiny_data |> 
  mutate(
    map_label = case_when(
      !is.na(preferred_species_for_reporting) ~ preferred_species_for_reporting,
      # Fall back to lowest classified taxonomy rank 
      Genus != "" & !is.na(Genus) ~ Genus,
      Family != "" & !is.na(Family) ~ Family,
      Order != "" & !is.na(Order) ~ Order,
      Class != "" & !is.na(Class) ~ Class,
      Phylum != "" & !is.na(Phylum) ~ Phylum,
      TRUE ~ Domain
      )
  )  |> 
  mutate(date = as.Date(sampling.date, format = "%m.%d.%y")) # Convert dates from character --> date
# 



```

Add common names - using indicator list and filling in NA's manually

```{r}
# Unique taxa - match species
taxa_list <- unique(shiny_data$map_label) |> 
  as.data.frame() |> 
  rename(map_label = 1) |>   # Rename first column
  filter(grepl(" ", map_label)) |> 
  left_join(indicators, by = c("map_label" = "scientific_name")) |> 
  select(map_label, common_name) |> 
  distinct(map_label, .keep_all = TRUE)

# Fill in those without common name match
taxa_list <- taxa_list %>%
  mutate(common_name = case_when(
    map_label == "Alopias vulpinus" ~ "Common thresher shark",
    map_label == "Citharichthys stigmaeus" ~ "Speckled sanddab",
    map_label == "Citharichthys xanthostigma" ~ "Longfin sanddab",
    map_label == "Diaphus theta" ~ "California headlightfish",
    map_label == "Hydrolagus colliei" ~ "Spotted ratfish",
    map_label == "Isurus oxyrinchus" ~ "Shortfin mako shark",
    map_label == "Lamna ditropis" ~ "Salmon shark",
    map_label == "Lampanyctus tenuiformis" ~ "Broadfin lampfish",
    map_label == "Lyopsetta exilis" ~ "Slender sole",
    map_label == "Mirounga angustirostris" ~ "Northern elephant seal",
    map_label == "Myliobatis californica" ~ "Bat ray",
    map_label == "Nannobrachium ritteri" ~ "Broadfin lampfish",
    map_label == "Paralichthys californicus" ~ "California halibut",
    map_label == "Pleuronichthys decurrens" ~ "Curlfin sole",
    map_label == "Protomyctophum crockeri" ~ "California flashlightfish",
    map_label == "Pseudobatos productus" ~ "Shovelnose guitarfish",
    map_label == "Sebastes entomelas" ~ "Widow rockfish",
    map_label == "Sebastes jordani" ~ "Shortbelly rockfish",
    map_label == "Sebastes saxicola" ~ "Stripetail rockfish",
    map_label == "Squalus suckleyi" ~ "Pacific spiny dogfish",
    map_label == "Stenobrachius leucopsarus" ~ "Northern lampfish",
    map_label == "Symbolophorus californiensis" ~ "California lanternfish",
    map_label == "Symphurus atricaudus" ~ "California tonguefish",
    map_label == "Tarletonbeania crenularis" ~ "Blue lanternfish",
    map_label == "Triakis semifasciata" ~ "Leopard shark",
    TRUE ~ common_name  # Keep existing value if not listed above
  ))

# Join with shiny data
shiny_data <- shiny_data |> 
  select(-common_name_final) |> 
  left_join(taxa_list, by = "map_label")

na_shiny <- shiny_data |> 
  filter(is.na(common_name)) |> 
  distinct(map_label)


# Fill in remaining NA's in shiny data (higher classification)
shiny_data <- shiny_data %>%
  mutate(common_name = case_when(
    map_label == "Sebastes" ~ "Rockfish",  # matches "All rockfish" or "Rockfish"
    map_label == "Pleuronectidae" ~ "Righteye flounders",  # family of flatfish
    map_label == "Rajidae" ~ "Skates",  # part of "All sharks and rays"
    map_label == "Diaphus" ~ "Lanternfish",  # genus of lanternfish -- CHECK
    map_label == "Pleuronectiformes" ~ "Flatfish",  # matches "Flatfish"
    map_label == "Paralichthys" ~ "Flounders",  # genus of flatfish
    map_label == "Cyclopsettidae" ~ "Sand whiffs",  # family of flatfish
    map_label == "Stenobrachius" ~ "Lampfish",  # genus within "Lanternfish"
    TRUE ~ common_name  # Keep existing value
  ))


na_shiny <- shiny_data |> 
  filter(is.na(common_name)) |> 
  distinct(map_label)

# update map label to have scientific name (common name), also update season to have year

shiny_data <- shiny_data |> 
   mutate(
    map_label = if_else(
      !str_detect(map_label, "\\("),  # Only format if not already formatted
      paste0(map_label, " (", common_name, ")"),
      map_label
    )
  ) |> 
  mutate(
    date = as.Date(sampling.date, format = "%m.%d.%Y"),  # Convert to date format if not already
    year = year(date),  # Extract year
    Season = paste(Season, year)  # Combine season and year
  )


# Fix station name and remove air samples
shiny_data <- shiny_data |>
  mutate(Station = if_else(
    sample_id == "20250724-M1-C-torp-B",  
    "C1",
    Station
  )) |> 
  filter(medium != "air")

# Change column names to be same as previous metadata sheet
shiny_data <- shiny_data |> 
  rename(
         temperature = ctd_temp,
         salinity = ctd_sal,
         oxygen = ctd_oxygen2)

# Save as an RDS to use in shiny app
saveRDS(shiny_data, here("shiny_dashboard", "data", "processed", "shiny_data.rds"))


  
```

# Preprocess bathymetry data

```{r}
# Bathymetry preprocessing

# Get start and end points for bathymetry 
start_lon <- -121.7796
start_lat <- 36.8117
end_lon <- -122.5352
end_lat <- 36.6775


# Add a small buffer so the points aren't on the edge
lon_min <- min(c(start_lon, end_lon)) - 0.1
lon_max <- max(c(start_lon, end_lon)) + 0.1
lat_min <- min(c(start_lat, end_lat)) - 0.1
lat_max <- max(c(start_lat, end_lat)) + 0.1

# Download the background bathymetry map using the bounding box
monterey_map <- getNOAA.bathy(lon1 = lon_min, 
                              lon2 = lon_max,
                              lat1 = lat_min, 
                              lat2 = lat_max, 
                              resolution = 1)

# Get transect
transect_data <- get.transect(monterey_map,
                              x1 = start_lon,
                              y1 = start_lat,
                              x2 = end_lon,
                              y2 = end_lat,
                              distance = TRUE)


saveRDS(monterey_map, here("shiny_dashboard", "data", "processed", "monterey_map.rds"))
saveRDS(transect_data, here("shiny_dashboard", "data", "processed", "transect_data.rds"))


 # Calculate distance along transect for each station

stations_temp <- tibble(
  Station = c("ES", "C1", "M1", "MARS", "M2", "OSWS"),
  Latitude = c(36.8117, 36.8045, 36.7502, 36.721, 36.7015, 36.6775),
  Longitude = c(-121.7796, -121.858, -122.0495, -122.1903, -122.3743, -122.5352)
) |> 
  arrange(Longitude) # Order west to east
  

  station_distances <- data.frame(
    station = character(),
    distance = numeric(),
    stringsAsFactors = FALSE
  )
  
  for(i in 1:nrow(stations_temp)) {
    # Get transect to each station from start point
    temp_transect <- get.transect(monterey_map,
                                  x1 = start_lon,
                                  y1 = start_lat,
                                  x2 = stations_temp$Longitude[i],
                                  y2 = stations_temp$Latitude[i],
                                  distance = TRUE)
    
    station_distances <- rbind(station_distances,
                               data.frame(station = stations_temp$Station[i],
                                          distance = max(temp_transect$dist)))
  }

saveRDS(station_distances, "shiny_dashboard/data/processed/station_distances.rds")
```
